### 2015-05-06 papers/histogram/thesis/thesis.tex
added subsection about end conditions, as well as some helpful figures

Due to the fact that Metropolis Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$; that is, the number of observations
$H\p{X_0}$ of the system with $X\p{s}=X_0$ is proportional to the
total number of states $s$ for which $X\p{s}=X_0$. It is sometimes the
case, however, that the density of states in some region $R$ in the
range of possible $X$ is so low that it is practically impossible (via
Metropolis Monte Carlo) to sufficiently sample $R$, that is, to
accumulate a statistically significant histogram $H\p{X\in R}$, in any
reasonable amount of time. Given that the entropy $S\p{X}$ can be
expressed in terms of the number of microstates $\Omega\p{X}$ as
$S\p{X}=k\ln\Omega\p{X}\propto\ln D\p{X}$, we may refer to regions $R$
low state densities $D\p{X\in R}$ as low entropy states.

@@

Histogram methods provide a means to address Metropolis Monte Carlo's
inability to sufficiently sample low entropy states by introducing a
bias into the otherwise random sampling of state space. A weighting
function $w$ is introduced, whose domain is the value of some property
$X\p{s}$ which depends on the system state $s$. An additional
condition is then added to step \ref{alg:metropolis_move} of Algorithm
\ref{alg:metropolis} in order to accept an attempted move: the weights
$w\sp{X\p{s_i}}$ and $w\sp{X\p{s_f}}$ of the initial (pre-move) state
$s_i$ and final (post-move) state $s_f$ are used to determine the
probability $P_m\p{s_i\to s_f}$ of accepting an otherwise valid move
via
\begin{align}
  P_m\p{s_i\to s_f}=\min\set{\f{w\sp{X\p{s_f}}}{w\sp{X\p{s_i}}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\sp{X\p{s_f}}>w\sp{X\p{s_i}}$, the move
$s_i\to s_f$ is accepted; when $w\sp{X\p{s_f}}<w\sp{X\p{s_i}}$, the
ratio of these weights determines the probability of accepting the
move. Due to the fact that only ratios of weights determine
$P_m\p{s_i\to s_f}$, the weights $w\sp{X\p{s}}$ are scale-invariant,
meaning that their effect on simulations is unchanged by scale factors
that are constant with respect to $X\p{s}$.

@@

In this paper, we will consider weights $w\p{E}$ which depend only on
the energy $E$ of the square-well fluid. Due to the fact that the
square-well fluid can only have discrete energies $E=-n\epsilon$,
where $n$ is a non-negative integer and $\epsilon$ is the well depth,
in simulation we store the weight function as an array of values. We
may therefore use the terms ``weight function,'' and ``weight array,''
and ``weights'' interchangeably.

\fixme{mention that simulating with flat weights $w\p{E}=w_0$ is the
  same as running a Metropolis MC simulation.}

@@

The partition function at $T=T_0$ for simulations with canonical
weights is
\begin{align}
  \tilde Z\p{T_0}=\sum_E\tilde D\p{E}e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{w\p{E}}~e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{e^{-E/kT_0}}~e^{-E/kT_0}=\sum_EH\p{E},
  \label{Z_canonical}
\end{align}
and the value of a thermodynamic property $X\p{T_0}$
\begin{align}
  X\p{T_0}=\f1{\tilde Z\p{T_0}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT_0}
  =\f{\sum_E\bk{X}_EH\p{E}}{\sum_EH\p{E}}.
  \label{eq:X_canonical}
\end{align}
The simplifications in in (\ref{Z_canonical}) and
(\ref{eq:X_canonical}), which have no explicit dependence on $T_0$,
occur because canonical Monte Carlo simulations sample energies in
proportion to the distribution (over energy) of microstates at a
temperature of $T_0$.  Canonical Monte Carlo simulations can therefore
fail to sufficiently sample energies which are important
(i.e. energies with a non-negligible state probability density) at
different temperatures. As a consequence, such simulations should not
be used to determine properties $X\p{T\ne T_0}$, and are thus referred
to as ``fixed temperature'' simulations.

Though canonical Monte Carlo is simple to implement, its inability to
investigate a system at more than one temperature at a time is a
disadvantage for determining the temperature dependence of system
properties. In order to find the behavior of $\bk{X}_T$, one must run
many simulations at discrete temperature intervals; each such
simulation will yield one sample (e.g. one data point) of $\bk{X}_T$.

Sampling low energies, however, or understanding system behavior at
low temperatures, is even more problematic with canonical
weights. Using low temperature canonical weights will indeed force a
simulated system down to low energies, but will also likely freeze the
system into a local minimum of its energy landscape. Freezing into a
state means that a simulation will sample only a small portion of the
energy landscape, even though there may (and generally will) be many
other states with the same energy.

Due to these problems, we will not use canonical weights alone to
study the square-well fluid. We will, however, use canonical weights
for part of all weight arrays, as discussed in Section
\ref{sec:min_energy}.

@@

Algorithm \ref{alg:flat_histogram} provides a naive way to implement
the flat histogram method without knowing a priori the density of
states $D\p{E}$. This algorithm involves first running a simulation
without weights (or with constant weights $w\p{E}=w_0$) while
collecting a histogram $H\p{E}$ of the observed energies $E$. After
some time (loosely defined), the histogram $H\p{E}$ may be taken as an
approximation of the density of states $\tilde D\p{E}$ (see Section
\ref{sec:histogram_methods}), and used via (\ref{eq:flat_weights}) to
generate a weight array for an actual simulation. Algorithm
\ref{alg:flat_histogram} has one free parameter, which determines the
amount of time for which to collect an energy histogram $H\p{E}$.

The problem with this implementation of the flat histogram method is
that histogram methods are generally necessary when some range of
energies $R$ is inaccessible via Metropolis Monte Carlo
simulations. In such a case, the energy histogram $H\p{E\in R}$ after
simulating for any reasonable amount of time will be statistically
insignificant or null, invalidating the approximation
$H\p{E}\approx\tilde D\p{E}$. This method will therefore only
``flatten'' the histogram at energies $E$ which have been sufficiently
sampled for $H\p{E}$ to be statistically significant. The end goal in
implementing histogram methods, however, is precisely to sample those
energies that cannot be sufficiently sampled without employing clever
schemes to bias Monte Carlo simulations. The flat histogram method
therefore does not directly help us study the square-well fluid. This
method does, however, motivate the method in Section
\ref{sec:simple_flat}, and provides some of the theory behind the
methods in Sections \ref{sec:tmmc} and \ref{sec:oetmmc}.

@@

\begin{algorithm}[!t]
  \caption{The simple flat method}
  \label{alg:simple_flat}
  \begin{alg}

@@

\begin{algorithm}[!b]
  \caption{Wang-Landau initialization of weights}
  \label{alg:wang_landau}
  \begin{alg}

@@

  \item If the histogram $H\p{E}$ is not sufficiently flat, i.e. if it
    fails a ``flatness'' condition $C\sp{H}$, return to step
    \ref{alg:wang_landau_sim}, otherwise
    \begin{enumerate}
    \item reset $H\p{E}=0$ for all $E$,
    \item increment $k\leftarrow k+1$,
    \item set $f_k=\sqrt[u]{f_{k-1}}$ for some predetermined $u>1$,
      \label{alg:wang_landau_update}
    \item if $f_k\ge c$ for some predetermined $c$, return to step
      \ref{alg:wang_landau_sim}.
    \end{enumerate}

@@

Algorithm \ref{alg:wang_landau} spells out the Wang-Landau
initialization process, which has five free parameters: the number of
iterations $n$ to run in step \ref{alg:wang_landau_sim}, the initial
factor $f_0$, the update factor $u$, the flatness condition $C\sp{H}$,
and the cutoff $c$. The first of these parameters, $n$, controls how
often to check the flatness condition $C\sp{H}$ and run updates, if
necessary. The value of $n$ is not too important, so long as it is
reasonable for the system at hand. Too small value of $n$ will waste
computation time determining $C\sp{H}$, while too large of a value
will waste time simulating when $C\sp{H}$ has already been
satisfied. In general, $n$ should be proportional to the computation
time of $C\sp{H}$, which scales with energy range of the simulated
system. We therefore use $n=L$, where $L$ is, as described in Section
\ref{sec:simple_flat}, the number of energy levels for the given
square-well fluid. Other works typically neglect providing any
heuristic or formula for $n$, reporting a system-independent
$n=10^3$\cite{wang_landau_mod} or $10^5$\cite{wang_landau}.
\fixme{find more citations for this last claim}

@@

The final free parameter in the Wang-Landau method is the cutoff $c$,
which determines how small $f$ can get before the algorithm stops
modifying the weight array. Too large of a value for $c$ will make
Wang-Landau quit prematurely, whereas too small a value will cause the
algorithm to waste time simulating after the weights $w\p{E}$ have
essentially converged. One might imagine using an alternate end
condition for Wang-Landau, which involves comparing the current
weights $w_k\p{E}$ to those at the end of the previous cycle,
$w_{k-1}\p{E}$, to check whether the initialization process is still
making appreciable modifications to the weights, but such a check
would require both a metric and a free parameter to define what an
appreciable change is. For simplicity, we stuck with the standard end
condition given in Algorithm \ref{alg:wang_landau}. As with the other
free parameters in Wang-Landau, we are not aware of any heuristics for
determining an appropriate value of $c$, but literature cites typical
values of $\ln c=10^{-8}$; we used $\ln c=10^{-6}$.

\subsection{The transition matrix Monte Carlo method}
\label{sec:tmmc}

\fixme{add citations}

Unlike the previous histogram methods, the transition matrix Monte
Carlo (TMMC) method does not use weights, and merely compute the
density of states $D\p{E}$ of a system\cite{tmmc}. The density of
states can in turn be used to determine flat histogram weights via
(\ref{eq:flat_weights}). TMMC introduces a new object: the energy
transition matrix, $T$, whose components $T_{ij}$ are the
probabilities that a system will transition {\it to} a state with
energy $E_i$ {\it from} a given state with energy $E_j$,
i.e. $T_{ij}=P\p{E_j\to E_i}$.

@@

Knowing the transition matrix of a system determines, among other
properties, the system's density of states. Consider an ensemble of
Metropolis Monte Carlo simulations with a distribution of energies at
a time $t$ given by $\hat D^{t}\p{E}$, such that the probability of
any given simulation to have an energy $E_i$ at a time $t$ is $\hat
D_i^{t}=\hat D^{t}\p{E_i}$. The probability $\hat D_i^{t}$ can be
expressed in terms of the distribution $\hat D^{t-1}\p{E}$ one time
step (i.e. one Metropolis Monte Carlo move) prior by
\begin{align}
  \hat D_i^{t}=\sum_jP\p{E_i\to E_j}\hat D_j^{t-1} =\sum_j T_{ij}\hat
  D_j^{t-1}.
  \label{eq:transition_evolution}
\end{align}
At equilibrium, the probability distributions $\hat D^{t}=\hat
D^{t-1}$. Furthermore, the distribution of states in an equilibrium
ensemble of Metropolis Monte Carlo simulations is precisely the
density of states $D\p{E}$\fixme{explain why?}, which means
$D_i=T_{ij}D_j$. Simultaneously expressing this condition for all
energies,
\begin{align}
  D=TD. \label{eq:dos_eigen}
\end{align}
Finding the density of states of a system can thus be reduced to
finding the eigenvector of the transition matrix which has a
corresponding unit eigenvalue\fixme{why is this eigenvector
  unique?}. Once one has determined the density of states of a system,
one can initialize weights via the flat histogram method,
i.e. (\ref{eq:flat_weights}).

Computationally determining the transition matrix of a system is
straightforward: one need only simulate the system and collect a
histogram $\tilde T_{ij}=\tilde T\p{E_i,E_j}$ of the energy
transitions after each Metropolis Monte Carlo move, i.e. after step
\ref{alg:metropolis_move} of Algorithm \ref{alg:metropolis}. A
histogram of transitions $\tilde T\p{E_i,\bar E_j}_{}$ from a fixed
energy $\bar E_j$ to energies $E_i$ is then proportional to the
probability distribution of transitions from $\bar E_j$, i.e. the
transition matrix $T\p{E_i,\bar E_j}$. Crucially, this proportionality
does not depend on the history of a simulation, or on how the
simulation got to be in the energy $\bar E_j$, so long as samples of
the energy $\bar E_j$ are themselves void of systemic
bias. Furthermore, one does not actually need to transition from $E_j$
to $E_i$ in order to collect statistics on $T\p{E_i,E_j}$: after
deciding whether to accept a move based on whether it results in a
valid system state, one can increment $\tilde T\p{E_i,E_j}$ before
deciding to reject the move for some other reason, e.g. because
$w\p{E_j}\gg w\p{E_i}$, or even because one wishes to collect more
statistics on $T\p{E_i,\bar E_j}$.

@@

While one could imagine many initialization routines to determine the
transition matrix $T$, we use Algorithm \ref{alg:tmmc}, which is a
modified version of a routine provided in \cite{tmmc_prl}\fixme{is
  this citation correct?}. The primary motivation behind this
algorithm is such: as lower energies are always more difficult to
sample than higher energies, transitions to lower energies should
always be accepted, while transitions to higher energies should pass a
probabilistic acceptance test. To understand the probability in step
\ref{alg:tmmc_prob} of Algorithm \ref{alg:tmmc} of accepting a move
$s_i\to s_f$ for which $E_f>E_i$, first consider the case of $c=0$ and
$T_{\t{min}}\to0$, in which case the probability becomes
\begin{align}
  P_m\p{s_i\to s_f}=\f{\p{\tilde T_d\p{E_i,\Delta E}}\big/
    \p{\sum_{\Delta E'}\tilde T_d\p{E_i,\Delta E'}}}
  {\p{\tilde T_d\p{E_f,-\Delta E}}\big/
    \p{\sum_{\Delta E'}\tilde T_d\p{E_f,\Delta E'}}}
  =\f{T\p{E_i,\Delta E}}{T\p{E_f,-\Delta E}}
  =\f{P\p{E_i\to E_f}}{P\p{E_f\to E_i}},
  \label{eq:tmmc_prob_simplified}
\end{align}
where $\Delta E=E_f-E_i$ and $P\p{E_m\to E_n}$ is the unbiased
probability of a transition from an energy $E_m$ to an energy
$E_n$. The result in (\ref{eq:tmmc_prob_simplified}) is the
probability of accepting moves $s_i\to s_f$ required for simulation
via Algorithm \ref{alg:tmmc} to yield a flat energy
histogram\cite{tmmc}. Unlike the previous histogram methods, we desire
a flat histogram during initialization via Algorithm \ref{alg:tmmc}
not because we are constructing weights $w\p{E}$, but because we wish
to sample the transition histogram $\tilde T_d\p{E,\Delta E}$ equally
at all energies $E$. \fixme{briefly explain
  (\ref{eq:tmmc_prob_simplified}) by considering a two-level system,
  and extending to larger systems}.

\begin{algorithm}[tb]
  \caption{Transition matrix Monte Carlo initialization}
  \label{alg:tmmc}
  \begin{alg}

@@

  \item If some predetermined initialization end condition has
    \emph{not} been met, return to step \ref{alg:tmmc_move}.

@@

\subsection{The optimized ensemble method}
\label{sec:optimized_ensemble}

All histogram methods introduced in this paper thus far have focused
on determining a weight array $w\p{E}$ for a flat energy histogram
$H\p{E}$, in order to sample and collect statistics on all energies
equally. This motivation considers the quantity of statistics, but not
their quality. Biased simulations collect statistics on low energy
states by first getting into these states, and then rejecting moves
which move the system into higher energy states. Statistics on low
energy states will therefore generally be highly correlated, as they
are based on many samples of only a few low energy regions of a
system's energy landscape. The optimized ensemble
(OE)\cite{optimized_ensemble} attempts to address the autocorrelation
between low energy samples by finding weights to maximize the rate at
which a simulation makes round trips between low and high energy
states. Maximizing the round trip rate, in turn, maximizes the rate at
which a simulation makes independent, uncorrelated samples of low
energies.

@@

The algorithm provided in \cite{optimized_ensemble} to determine and
construct weights given by \ref{eq:optimized_weights} involves
simulating with and iteratively modifying flat histogram weights. This
algorithm is not meant to construct weights from scratch, but rather
to optimize them. We therefore will not be comparing the optimized
ensemble to other methods. We do, however, credit this method for
motivating the hybrid OE-TMMC method in Section \ref{sec:oetmmc}.

@@

\subsubsection{State of maximal entropy}
\label{sec:max_entropy}

@@

\subsubsection{Minimum important energy}
\label{sec:min_energy}

@@

\subsubsection{Computing the density of states}
\label{sec:dos}

\fixme{This section might just become a part of the section on the
  minimum important energy. (\ref{eq:dos}) gives us an explicit
  formula for the density of states if we are confident in our energy
  histogram. We can also compute the density of states from the
  transition matrix, which is sometimes more appropriate to do.}

\subsection{End conditions}
\label{sec:end_conditions}

% \fixme{Talk about how we want method-independent end conditions, if
% possible. Introduce the concept of statistically independent energy
% samples, and explain two measures (one optimistic, and one
% pessimistic) for determining the number of statistically independent
% times we have samples a given energy.}

>>>>

Due to the fact that Metropolis Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$; that is, the number of observations
$H\p{X_0}$ of the system with $X\p{s}=X_0$ is proportional to the
total number of states $s$ for which $X\p{s}=X_0$. It is sometimes the
case, however, that the density of states in some range $R$ of
possible $X$ is so low that it is practically impossible (via
Metropolis Monte Carlo) to sufficiently sample $R$, that is, to
accumulate a statistically significant histogram $H\p{X\in R}$, in any
reasonable amount of time. Given that the entropy $S\p{X}$ can be
expressed in terms of the number of microstates $\Omega\p{X}$ as
$S\p{X}=k\ln\Omega\p{X}\propto\ln D\p{X}$, we may refer to regions
with low state densities $D\p{X}$ as low entropy states.

Figure \ref{fig:dos_sample} provides an example of an unnormalized
density of states $\tilde D\p{E}$ in energy for a particular
square-well fluid. The density of states has been ``normalized'' to
have a maximum value of $1$ in order to emphasize the logarithmic span
of $D\p{E}$. This logarithmic span ($\sim10^{45}$) is proportional to
the number of moves which one must simulate via Metropolis Monte Carlo
in order to observe the energies with the lowest state densities. For
reference, the world's current fastest supercomputers can perform
$\sim10^{16}$ floating-point operations per second.

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/dos-sample.pdf}
  \caption[Sample density of states]{Sample density of states for
    square-well fluid with a well width $\lambda=1.3$, filling
    fraction $\eta=0.3$, and $N=20$ spheres, as computed by several
    histogram methods. The density of states has been normalized to
    have a maximum value of $1$.}
  \label{fig:dos_sample}
\end{figure}

@@

Histogram methods provide a means to address Metropolis Monte Carlo's
inability to sufficiently sample low entropy states by introducing a
bias into the otherwise random sampling of state space. A weighting
function $w$ is introduced, whose domain is the value of some property
$X\p{s}$ which depends on the system state $s$. An additional
condition is then added to step \ref{alg:metropolis_move} of Algorithm
\ref{alg:metropolis} in order to accept an attempted move: the weights
$w\p{X_i}$ and $w\p{X_f}$ of the initial (pre-move) state $s_i$ and
final (post-move) state $s_f$, respectively, are used to determine the
probability $P_m\p{s_i\to s_f}$ of accepting an otherwise valid move
via
\begin{align}
  P_m\p{s_i\to s_f}=\min\set{\f{w\p{X_f}}{w\p{X_i}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\p{X_f}>w\p{X_i}$, the move $s_i\to
s_f$ is accepted; when $w\p{X_f}<w\p{X_i}$, the ratio of these weights
determines the probability of accepting the move. Due to the fact that
only ratios of weights determine $P_m\p{s_i\to s_f}$, the weights
$w\p{X}$ are scale-invariant, meaning that their effect on simulations
is unchanged by scale factors that are constant with respect to
$X\p{s}$. Furthermore, simulating with flat weights $w\p{X}=w_0$ for
all $X$ is equivalent to simulating without weights, as the
probability $P_m\p{s_i\to s_f}$ will always equal 1.

@@

In this paper, we will use a histogram $H\p{E}$, weights $w\p{E}$, and
density of states $D\p{E}$ which are functions of the square-well
fluid's energy $E$. Due to the fact that the square-well fluid can
only have discrete energies $E=-n\epsilon$, where $n$ is a
non-negative integer and $\epsilon$ is the well depth, in simulation
we store the weight function as an array of values. We will therefore
generally refer to the weight function as a ``weight array,'' or
simply ``weights.''

@@

The partition function at $T=T_0$ for simulations with canonical
weights is
\begin{align}
  \tilde Z\p{T_0}=\sum_E\tilde D\p{E}e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{w\p{E}}~e^{-E/kT_0}
  =\sum_E\f{H\p{E}}{e^{-E/kT_0}}~e^{-E/kT_0}=\sum_EH\p{E},
  \label{Z_canonical}
\end{align}
and the value of a thermodynamic property $X\p{T_0}$
\begin{align}
  X\p{T_0}=\f1{\tilde Z\p{T_0}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT_0}
  =\f{\sum_E\bk{X}_EH\p{E}}{\sum_EH\p{E}},
  \label{eq:X_canonical}
\end{align}
where $\bk{X}_E$ is the mean value of $X$ at the energy $E$. The
simplifications in in (\ref{Z_canonical}) and (\ref{eq:X_canonical}),
which have no explicit dependence on $T_0$, occur because canonical
Monte Carlo simulations sample energies in proportion to the
distribution (over energy) of microstates at a temperature of $T_0$.
Canonical Monte Carlo simulations can therefore fail to sufficiently
sample energies which are important (i.e. energies with a
non-negligible state probability density) at different
temperatures. As a consequence, such simulations should not be used to
determine properties $X\p{T\ne T_0}$, and are thus referred to as
``fixed temperature'' simulations.

Though canonical Monte Carlo is simple to implement, its inability to
investigate a system at more than one temperature at a time is a
disadvantage for determining the temperature dependence of system
properties. In order to find the behavior of $\bk{X}_T$, one must run
many simulations at discrete temperature intervals; each such
simulation will yield one data point on $\bk{X}_T$.

Sampling low energies, however, or understanding system behavior at
low temperatures, is even more problematic with canonical
weights. Using low temperature canonical weights will indeed force a
simulated system down to low energies, but will also likely freeze the
system into a local minimum of its energy landscape. Freezing into a
state in this manner will limit a simulation to sampling only a small
portion of the energy landscape, even though there may (and generally
will) be many other states with the same energy.

Due to these problems, we will not use canonical weights alone to
study the square-well fluid. We will, however, use canonical weights
for portions of all weight arrays, which we discuss in Section
\ref{sec:energy_range}.

@@

Algorithm \ref{alg:flat_histogram} provides a naive way to implement
the flat histogram method without knowing a priori the density of
states $D\p{E}$. This algorithm involves first running a simulation
without weights (or with constant weights $w\p{E}=1$) while collecting
a histogram $H\p{E}$ of the observed energies $E$. After some time
(loosely defined), the histogram $H\p{E}$ may be taken as an
approximation of the density of states $\tilde D\p{E}$, and used via
(\ref{eq:flat_weights}) to generate a weight array for an actual
simulation. Algorithm \ref{alg:flat_histogram} has one free parameter,
which determines the amount of time for which to collect an energy
histogram $H\p{E}$.

The problem with this implementation of the flat histogram method is
that histogram methods are generally necessary when some range of
energies $R$ is inaccessible via Metropolis Monte Carlo
simulations. In such a case, the energy histogram $H\p{E\in R}$ after
simulating for any reasonable amount of time will be statistically
insignificant or null, invalidating the proportionality $H\p{E}\propto
D\p{E}$. This method will therefore only ``flatten'' the histogram at
energies $E$ which have been sufficiently sampled for $H\p{E}$ to be
statistically significant. The end goal in implementing histogram
methods, however, is precisely to sample those energies that cannot be
sufficiently sampled without employing clever schemes to bias Monte
Carlo simulations. Therefore, this method does not directly help us
study the square-well fluid. This method does, however, motivate the
method in Section \ref{sec:simple_flat}, and provides some of the
theory behind the methods in Sections \ref{sec:tmmc} and
\ref{sec:oetmmc}.

@@

\begin{algorithm}[!b]
  \caption{The simple flat method}
  \label{alg:simple_flat}
  \begin{alg}

@@

\begin{algorithm}[tb]
  \caption{Wang-Landau initialization of weights}
  \label{alg:wang_landau}
  \begin{alg}

@@

  \item If the histogram $H\p{E}$ is not sufficiently flat, i.e. if it
    fails a ``flatness'' condition $C\sp{H}$, return to step
    \ref{alg:wang_landau_sim}, otherwise
    \begin{enumerate}
    \item reset $H\p{E}=0$ for all $E$,
    \item increment $k\leftarrow k+1$,
    \item set $f_k=\sqrt[u]{f_{k-1}}$ for some predetermined $u>1$,
      \label{alg:wang_landau_update}
    \item if $f_k\ge c$ for some predetermined $c$, return to step
      \ref{alg:wang_landau_sim}.
      \label{alg:wang_landau_end}
    \end{enumerate}

@@

Algorithm \ref{alg:wang_landau} spells out the Wang-Landau
method. This algorithm includes four free parameters which affect
initialization process: the number of iterations $n$ to run in step
\ref{alg:wang_landau_sim}, the initial factor $f_0$, the update factor
$u$, and the flatness condition $C\sp{H}$. The cutoff $c$ determines
the condition for Algorithm \ref{alg:wang_landau} to terminate.

The first of these parameters, $n$, controls how often to check the
flatness condition $C\sp{H}$ and run updates, if necessary. The value
of $n$ is not too important, so long as it is reasonable for the
system at hand. Too small value of $n$ will waste computation time
determining $C\sp{H}$, while too large of a value will waste time
simulating when $C\sp{H}$ has already been satisfied. In general, $n$
should be proportional to the computation time of $C\sp{H}$, which
scales with energy range of the simulated system. We therefore use
$n=L$, where $L$ is, as described in Section \ref{sec:simple_flat},
the number of energy levels for the given square-well fluid. Other
works typically neglect providing any heuristic or formula for $n$,
reporting a system-independent $n=10^3$\cite{wang_landau_mod} or
$10^5$\cite{wang_landau}.  \fixme{find more citations for this last
  claim}

@@

The final free parameter in the Wang-Landau method is the cutoff $c$,
which determines how small $f$ can get before the algorithm stops
modifying the weight array. Too large of a value for $c$ will make
Wang-Landau quit prematurely, whereas too small a value will cause the
algorithm to waste time simulating after the weights $w\p{E}$ have
essentially converged. One might imagine using an alternate end
condition for Wang-Landau, which involves comparing the current
weights $w_k\p{E}$ to those at the end of the previous cycle,
$w_{k-1}\p{E}$, to check whether the initialization process is still
making appreciable modifications to the weights, but such a check
would require both a metric and a free parameter to define what an
appreciable change is. For simplicity, we stuck with the standard end
condition given in Algorithm \ref{alg:wang_landau}. As with the other
free parameters in Wang-Landau, we are not aware of any heuristics for
determining an appropriate value of $c$, but literature cites typical
values of $\ln c=10^{-8}$. In order to compare different histogram
methods fairly, we used an alternate end condition in step
\ref{alg:wang_landau_end} of Algorithm \ref{alg:wang_landau}
(discussed in Section \ref{sec:end_conditions}), and therefore did not
need to choose a value of $c$.

\subsection{The transition matrix Monte Carlo (TMMC) method}
\label{sec:tmmc}

Unlike the previous histogram methods, the transition matrix Monte
Carlo method does not use weights, and merely compute the density of
states $D\p{E}$ of a system\cite{tmmc}. The density of states can in
turn be used to determine flat histogram weights via
(\ref{eq:flat_weights}). TMMC introduces a new object: the energy
transition matrix, $T$, whose components $T_{ij}$ are the
probabilities that a system will transition {\it to} a state with
energy $E_i$ {\it from} a given state with energy $E_j$,
i.e. $T_{ij}=P\p{E_j\to E_i}$.

@@

Knowing the transition matrix of a system determines, among other
properties, the system's density of states. Consider an ensemble of
Metropolis Monte Carlo simulations with a distribution of energies at
a time $t$ given by $\hat D^{t}\p{E}$, such that the probability of
any given simulation to have an energy $E_i$ at a time $t$ is $\hat
D_i^{t}=\hat D^{t}\p{E_i}$. The probability $\hat D_i^{t}$ can be
expressed in terms of the distribution $\hat D^{t-1}\p{E}$ one time
step (i.e. one Metropolis Monte Carlo move) prior by
\begin{align}
  \hat D_i^{t}=\sum_jP\p{E_i\to E_j}\hat D_j^{t-1} =\sum_j T_{ij}\hat
  D_j^{t-1}.
  \label{eq:transition_evolution}
\end{align}
At equilibrium, the probability distributions do not vary with time,
meaning $\hat D^{t}=\hat D^{t-1}$. Furthermore, the distribution of
states in an equilibrium ensemble of Metropolis Monte Carlo
simulations is precisely the density of states $D\p{E}$, which means
$D_i=T_{ij}D_j$.  Simultaneously expressing this condition for all
energies,
\begin{align}
  D=TD. \label{eq:dos_eigen}
\end{align}
Finding the density of states of a system can thus be reduced to
finding the eigenvector of the transition matrix which has a
corresponding unit eigenvalue\fixme{why is this eigenvector
  unique?}. Once one has determined the density of states of a system,
one can initialize weights via the flat histogram method,
i.e. (\ref{eq:flat_weights}).

Computationally determining the transition matrix of a system is
straightforward: one need only simulate the system and collect a
histogram $\tilde T_{ij}=\tilde T\p{E_i,E_j}$ of the energy
transitions after each Metropolis Monte Carlo move, i.e. after step
\ref{alg:metropolis_move} of Algorithm \ref{alg:metropolis}. A
histogram of transitions $\tilde T\p{E_i,\bar E_j}$ from a fixed
energy $\bar E_j$ to energies $E_i$ is then proportional to the
probability distribution of transitions from $\bar E_j$, i.e. the
transition matrix $T\p{E_i,\bar E_j}$. Crucially, this proportionality
does not depend on the history of a simulation, or on how the
simulation got to be in the energy $\bar E_j$, so long as samples of
the energy $\bar E_j$ are themselves void of systemic
bias. Furthermore, one does not actually need to transition from $E_j$
to $E_i$ in order to collect statistics on $T\p{E_i,E_j}$: after
deciding whether to accept a move based on whether it results in a
valid system state, one can increment $\tilde T\p{E_i,E_j}$ before
deciding to reject the move for some other reason, e.g. because
$w\p{E_j}\gg w\p{E_i}$, or even because one wishes to collect more
statistics on $T\p{E_i,\bar E_j}$.

@@

\begin{algorithm}[!b]
  \caption{Transition matrix Monte Carlo initialization}
  \label{alg:tmmc}
  \begin{alg}

@@

  \item If some predetermined initialization end condition has
    \emph{not} been met, return to step
    \ref{alg:tmmc_move}. Otherwise, compute the density of states
    $D\p{E}$ by solving (\ref{eq:dos_eigen}), and set weights
    $w\p{E}=1/D\p{E}$.

@@

Figure \ref{fig:transitions_sample} provides an example of a
transition matrix $T_d\p{E,\Delta E}$ for a particular square-well
fluid. The yellow band at $\Delta E=0$ indicates that most moves
transition from an energy to itself. Low energies are more likely to
transition up in energy, with $\Delta E>0$, than down in energy, with
$\Delta E<0$. The high energies which are more likely to transition
down in energy than up are those above the state of maximum entropy.

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/transitions-sample.pdf}
  \caption[Sample transition matrix]{Sample transition matrix
    $T_d\p{E,\Delta E}$ for square-well fluid with a well width
    $\lambda=1.3$, filling fraction $\eta=0.3$, and $N=20$ spheres.}
  \label{fig:transitions_sample}
\end{figure}

While one could imagine many initialization routines to determine the
transition matrix $T$, we use Algorithm \ref{alg:tmmc}, which is a
modified version of a routine provided in \cite{tmmc_prl}\fixme{is
  this citation correct?}. The primary motivation behind this
algorithm is such: as lower energies are always more difficult to
sample than higher energies, transitions to lower energies should
always be accepted, while transitions to higher energies should pass a
probabilistic acceptance test. To understand the probability in step
\ref{alg:tmmc_prob} of Algorithm \ref{alg:tmmc} of accepting a move
$s_i\to s_f$ for which $E_f>E_i$, first consider the case of $c=0$ and
$T_{\t{min}}\to0$, in which case the probability becomes
\begin{align}
  P_m\p{s_i\to s_f}=\f{\p{\tilde T_d\p{E_i,\Delta E}}\big/
    \p{\sum_{\Delta E'}\tilde T_d\p{E_i,\Delta E'}}} {\p{\tilde
      T_d\p{E_f,-\Delta E}}\big/ \p{\sum_{\Delta E'}\tilde
      T_d\p{E_f,\Delta E'}}} =\f{T_d\p{E_i,\Delta
      E}}{T_d\p{E_f,-\Delta E}} =\f{P\p{E_i\to E_f}}{P\p{E_f\to E_i}},
  \label{eq:tmmc_prob_simplified}
\end{align}
where $\Delta E=E_f-E_i$ and $P\p{E_m\to E_n}$ is the unbiased
probability of a transition from an energy $E_m$ to an energy
$E_n$. The result in (\ref{eq:tmmc_prob_simplified}) is the
probability of accepting moves $s_i\to s_f$ required for simulation
via Algorithm \ref{alg:tmmc} to yield a flat energy
histogram\cite{tmmc}. Unlike the previous histogram methods, we desire
a flat histogram during initialization via Algorithm \ref{alg:tmmc}
not because we are constructing weights $w\p{E}$, but because we wish
to sample the transition histogram $\tilde T_d\p{E,\Delta E}$ equally
at all energies $E$. \fixme{briefly explain
  (\ref{eq:tmmc_prob_simplified}) by considering a two-level system,
  and extending to larger systems}.

\subsection{The optimized ensemble (OE) method}
\label{sec:optimized_ensemble}

All histogram methods introduced in this paper thus far have focused
on determining a weight array $w\p{E}$ for a flat energy histogram
$H\p{E}$, in order to sample and collect statistics on all energies
equally. This motivation considers the quantity of statistics, but not
their quality. Biased simulations collect statistics on low energy
states by first getting into these states, and then rejecting moves
which move the system into higher energy states. Statistics on low
energy states will therefore generally be highly correlated, as they
are based on many samples of only a few low energy regions of a
system's energy landscape. The optimized
ensemble\cite{optimized_ensemble} attempts to address the
autocorrelation between low energy samples by finding weights to
maximize the rate at which a simulation makes round trips between low
and high energy states. Maximizing the round trip rate, in turn,
maximizes the rate at which a simulation makes independent,
uncorrelated samples of low energies.

@@

The algorithm provided in \cite{optimized_ensemble} to construct
weights given by (\ref{eq:optimized_weights}) finds the diffusivity
$\alpha$ by solving for it in
(\ref{eq:walker_current_definition}). This procedure accepts flat
histogram weights, or equivalently the density of states, as an input,
and is not meant to construct weights from scratch. We therefore will
not be comparing the optimized ensemble to other methods in this
paper. We do, however, credit \cite{optimized_ensemble} for motivating
the hybrid OE-TMMC method in Section \ref{sec:oetmmc}.

@@

@@

@@
